Measures of genetic diversification in somatic tissues at bulk and single cell resolution suggest sources of unknown stochasticity

Version 1

Abstract

Abstract

Intra-tissue genetic heterogeneity is universal to both healthy and cancerous tissues. It emerges from the stochastic accumulation of somatic mutations throughout development and homeostasis. By combining population genetics theory and genomic information, genetic heterogeneity can be exploited to infer tissue organisation and dynamics in vivo. However, many basic quantities, for example the dynamics of tissue specific stem cells remain difficult to quantify precisely. Here we show that single cell and bulk sequencing data inform on different aspects of the underlying stochastic processes. Bulk derived variant allele frequency spectra (VAF) show transitions from growing to constant stem cell populations with age in samples of healthy oesophagus epithelium. Single-cell mutational burden distributions allow a sample size independent measure of mutation and proliferation rates. Mutation rates in adult haematopietic stem cells are higher compared to inferences during development, suggesting additional proliferation independent effects. Furthermore, single cell derived VAF spectra contain information on the number of tissue specific stem cells. In haematopiesis, we find approximately 2 × 105 HSCs, if all stem cells divide symmetrically. However, the single cell mutational burden distribution is over-dispersed compared to a model of Poisson distributed random mutations suggesting currently unknown processes induce additional stochasticity. Possible candidates for these processes may be occasional bursts of stem cell divisions, potentially in response to injury, or non-constant mutation rates either through environmental exposures or cell intrinsic variation.

Introduction

Intra-tissue genetic heterogeneity emerges from a multitude of dynamical processes [1, 2]. Cells randomly accumulate mutations, they self-renew, differentiate or die and clones of varying fitness may compete and expand [3–6]. These core evolutionary principals form the basis for describing the aging of somatic tissues, tumour initiation and tumour progression [7–12]. Naturally, there is great interest in understanding precisely how these dynamical processes act. This typically involves hypothesizing models of cell behavior, and deriving quantitative estimates of their underlying physical parameters [3, 5, 13–17]. These could by quantities such as mutation rates per cell division, active stem cell numbers, symmetric and asymmetric division rates, etc.. Some such parameters are already well characterised in certain tissues. We know for example that during early development healthy somatic cells accumulate on average between 1.2 and 1.3 mutations per genome per division [5, 18]. Other parameters, such as the number of stem cells in a tissue type, are well known in some tissues but harder to quantify in others.

In general, a lack of time resolved information complicates evolutionary inferences, and we must often rely on indirect information, for example the patterns of tissue-specific genetic heterogeneity within and across individuals [19]. These observed patterns of genetic heterogeneity emerge from the underlying stochastic processes and can strongly depend on the measurement technique [20]. Both enforce specific limitations on the inferability of the underlying dynamics. Different types of genomic data – such as bulk and single cell whole genome sequencing – may contain different information on the system, and each comes with their own limitations. When bulk sequencing data is used for inference, some evolutionary parameters such as the population size and proliferation rates of stem cells are entangled and cannot be obtained in isolation [3, 16]. If on the other hand single cell data is employed, some additional information becomes available, for example the individual cell mutational burdens and the co-occurrence of mutations. However, often the sampling size is sparse compared to the presumed number of long-term proliferating cells in many tissues [18, 21, 22], which in turn introduces other constraints. Here we show that combining such data obtained from different resolutions (bulk or single cell) can help to overcome limitations and further narrow possible ranges of inferred parameters. Here, we develop concrete mathematical and computational models for extracting evolutionary parameters from bulk and single cell information and apply these methods to whole genome bulk sequencing data in healthy oesophagus [10] and whole genome single cell sequencing data in healthy haematopoiesis [18].

Results

A stochastic model of mutation accumulation in healthy somatic tissue

We model the stem cell dynamics of healthy tissue as a collection of individual cells that divide, differentiate, and die stochastically at predefined rates (see Fig.1). Novel mutations can occur with every cell division, each daughter acquiring a random number drawn from a Poisson distribution with rate µ. We explicitly include symmetric and asymmetric stem cell divisions at potentially different rates, the former resulting in two stem or two differentiated cells, and the latter in one stem and one differentiated cell. Since differentiated cells are lost from the stem cell pool they are functionally equivalent to cell death. While only symmetric divisions allow variants to change in frequency, both division types introduce novel variants into the population and thus both contribute to the evolving heterogeneity.

The distribution of variant allele frequencies changes with the growth phases and by sampling.

(a) In the current population, cells divide symmetrically into two daughter cells or asymmetrically with only one daughter cell kept in the focused population. All other events are mathematically equivalent to and are treated as a part of cell death. (b) The rates of symmetric and asymmetric division change during the population growth and lead to a dynamical distribution of variant allele frequencies. (c) The observed VAF distribution is shifted again during sampling compared to the VAF of the whole population – a fact should be considered when inferring population properties through genetic data.

Mutation accumulation occurs through all developmental stages from conception to death [23]. We therefore consider three phases of demographic changes of the stem cell population. In the first early developmental phase the stem cell population rapidly expands from a single cell exclusively by symmetric divisions at rate γ. This is followed by a growth and maintenance phase wherein the population continues to expand but also undergoes turnover through asymmetric divisions at rate ϕ, as well as cell removal (due to differentiation or death) and replacement at rate ρ. In the final mature phase, cell turnover continues but the size of the total stem population N remains constant.

We employ two independent implementations of our theoretical model. One is a direct stochastic simulation utilizing a Gillespie algorithm. This approach creates a simulated dataset containing all single cell or bulk sequencing properties, however it can be computationally expensive. In addition, we compute the time-dynamical expected value of the distribution of variant allele frequencies (the VAF spectrum) directly, which we find obeys the partial differential equation where κ = f N(t) denotes the number of cells sharing a variant (the variant frequency f times the total population size N), δ(x) is the Dirac impulse function, and

This set of equations allows for computationally efficient numerical solutions (SI 1.3).

Transition of developmental growth and constant population size signatures in bulk whole genome sequencing data of healthy adult oesophagus

We first discuss established properties of the VAF spectrum relating to somatic tissues. In certain model systems it is given by a power law ∝ f −α with a critical exponent α. The value of the exponent depends on the demographic dynamics of the population. For a well-mixed exponentially growing population without cell death the VAF spectrum v(f) is given by 2µ/(f + f 2) (a f −2 power law) and is independent of time [6]. In contrast, for a population of constant size the spectrum obeys v(f) ∝ 2µ/f [24] (a f −1 power law; see also SI 1.5.1), though this solution is only approached at sufficiently long times. In the following, we focus on the dynamics of genetic diversification in healthy tissues and show that the expected VAF spectrum contains both contributions of growing and constant population dynamics.

Healthy adult somatic tissues are thought to initially expand rapidly during ontogenic growth, continue to expand during infancy and childhood, and reach homeostasis by adulthood. It is thus natural to ask what the expected VAF spectrum would look like in tissues experiencing homeostasis after a period of growth. We first investigated the underlying VAF dynamics in a minimal theoretical model where the population switches from pure exponential growth to constant size once a maximal cell number is reached. Stochastic simulations and numerical solutions of Eq. (1) show that the VAF distribution exhibits a gradual transition from the f −2 to the f −1 power law (Fig. 2a). These transitional states themselves do not adhere to some intermediate power-law (e.g. f −α for 1 < α < 2), but instead present a wavelike front whereby the equilibrium establishes earliest at the lowest frequencies and moves to higher frequency over time. As the distribution approaches f −1, convergence towards equilibrium continues to slow down, resulting in a higher density of solutions for evenly-spaced observation times. The transition from an exponential signature f −2 to a constant signature f −1 is initially fast for low frequency variants and slows down considerably at older ages for high frequency states. More complex forms of such demographic changes, for example the inclusion of a mixed growth-and- maintenance phase or a logistically growing population, result in qualitatively similar behavior (see Supplementary Fig. 5).

Bulk sequencing based VAF and mutation rate inferences in healthy esophagus.

(a) Simulated VAF with an initial logarithmic growth phase and subsequent constant population phase at different time points. As the population gets closer to the maximum simulated carrying capacity of N = 104, the VAF moves from a 1/f 2 growing population shape to a 1/f constant population shape. Note, the shift considerably slows at older age. (b) VAF from healthy tissue in the oesophagus of 9 individuals sorted into age brackets. The youngest bracket, 20-39, is closer to the developmental 1/f 2 scaling. The older age brackets are both close to the constant population 1/f scaling, resembling the theoretical expectations. (c) Inferred mutation rates increase linearly with age. (d) Simulations of slowly growing stem cell populations reveal that mutation rates appear to increase with age, although the true underlying per division mutation rate remaining constant.

To test whether adult human tissues show such transitional signatures, we grouped the VAFs of healthy adult oesophagus samples from Martincorena and colleagues [10] into three age groups (young, middle and old, Fig. 2b). In accordance with our theoretical expectations, the averaged VAF spectrum of the young donor group is closest to the expected f −2 distribution of ontogenic growth, with low frequency variants only starting to approach the f −1 homeostatic scaling. The averaged spectra of the middle and old age groups on the other hand notably shift towards the expected f −1 homeostatic line. Interestingly, while there is a clear separation between the spectra of the youngest and middle age groups, those of the middle and older groups are much closer. This agrees with our prediction that the speed of convergence of the spectra towards homeostatic scaling slows down with age.

VAF based mutation rate inferences in healthy tissues

The theoretical prediction for the VAF spectrum of an exponentially growing population, which is recovered in many cancers given sufficient sequencing depth [20], contains information on the effective mutation rate µ [25]. Similarly, the VAF spectrum of a constant population includes the same mutation rate µ (SI). Thus we can in principle infer µ from the previously described oesophagus data, for example by applying a regression approach. However, since the scaling of the VAF spectrum of healthy tissues can be shown to be age dependent (Fig. 2a-b), we use two reference shapes for a growing and a constant population, respectively. The fitting is performed by a linear regression least-squares approach. The general solution is computed from the VAF spectrum of the data, Vd(µ), and the reference shape Vr, which assumes µ = 1, as follows:

If Vd(µ) has the shape of Vr, then E(Vd(µ)) = µ E(Vr), which can be used to prove equation 2:

The VAF shapes for the growing and constant population give close estimates, as can be seen in fig. 2, which compares the age of all individuals with corresponding inferred effective mutation rates. The estimates are in the range of 1 to 2 mutations per genome per cell division and agree with previous observations based on early developmental stem cell divisions [18]. Estimates based on a growing population are slightly higher compared to estimates based on a constant population, but differences are small.

Surprisingly, we observe a clear trend of a linearly increasing effective mutation rate with age. This trend cannot be explained by the transition of variants from a growing into a constant population alone. An actual increase of the true mutation rate per cell division with age would be a direct explanation. However, this seems unlikely: It is well established that the mutational burden across individuals and tissues increases linearly with age [9, 11, 26], and an increasing mutation rate would in contrast result in an accelerated increase of mutational burden with age, which is not observed experimentally.

An alternative explanation is a continued slow linear increase of the stem cell population with age (Fig. 2d). In such a scenario, the effective mutation rate would appear to increase linearly with age as well, although the true mutation rate per cell division remains unchanged. At first glance, a linearly increasing stem cell population appears unnatural. However, such a linear increase would emerge from a small bias of stem cells towards self-renewal (growth) and concomitantly a slowdown of the stem cell proliferation rate proportional to the change of population size [27]. The latter is a natural feedback mechanism ensuring a constant output of differentiated somatic cells. Direct in vivo imaging in multiple human tissues including Oesaphagus seem to support an increased density and slower proliferation of stem cells with age [28]. This explanation is attractive for another reason: A feedback of decreased cell proliferation with increasing population size would also function as a tumour suppressor mechanism, and may be another reason cancer driver mutations in healthy tissues are abundant but rarely progress to cancer [10, 29]. Progression would require additional genomic or environmental changes to overcome this regulatory feedback.

Single-cell mutational burden allows sample size independent inferences of stem cell proliferation and mutation rates

With the increasing availability of single-cell data, it becomes possible to investigate distributions of certain quantities for which bulk sequencing only provides average measurements. One such is the number of mutations present in individual cells. Recent work by Lee-Six and colleagues [18] showed that in a 59 year old donor haematopoietic stem cells had accumulated on average around 1000 mutations per cell; however there was significant variation between individual genomes, some carrying 900 mutations, while others up to 1200. This variation is the result of the stochastic nature of mutation accumulation and cell division, and can be exploited to estimate per division mutation and stem cell proliferation rates. To this end we look at the distribution of the number of mutations per cell, which we will from here on refer to as the mutational burden distribution. Formally, we describe the mutational burden mj in a cell j by a stochastic sum over its past divisions , in which both the total number of divisions yj in the cell’s past and the number of mutations per division uij are random variables. Even without knowing the actual distributions of y and u, general expressions for the expectation and variance of mj that only depend on the expectation and variance of y and u exist and are given by E(m) = E(y) E(u) and Var(m) = E(y) Var(u) + E(u)2 Var(y) (SI 1.7). These expressions allow direct estimates of the effective division and mutation rate. Assuming the mutation rate per daughter cell is Poisson distributed with expectation µ, the average number of divisions per lineage — expressed in terms of an effective division rate λ — becomes

In addition, if the mutation rate µ is unknown, it can similarly be obtained from

We first show that these estimators recover per division mutation and proliferation rates in stochastic simulations (Fig. 3a and b). More importantly, they only require information on comparably few single cells. Approximately, 100 cells are sufficient to reliably reconstruct the mutational burden distribution, which for example would constitute a small sample of the haematopoietic stem cell pool. In fact, the inference does not significantly improve if we increase data resolution beyond a few hundred single cells (Fig. 3a and b). This is in stark contrast to inferences based on single cell phylogeny: Although they contain more information in principle, important aspects of a single cell phylogeny, e.g. the number of observable branchings, are strongly affected by sampling and stem cell population size.

Inference of evolutionary parameters on simulated stem cell populations.

Simulated populations were run up to age 59, growing exponentially from a single cell to constant size NM = 10′000 at age tM = 5, with mutation rate µ = 1.2 and division rates λ = 5 and p = 0.4. Where sampling is mentioned, the sample size 89 was taken. a) The single-cell mutational burden distribution. The compound Poisson distribution (dashed line) matches the burden distribution when averaging over multiple independently evolved populations (filled curve). b) Distribution of estimated mutation rates from 10’000 individual simulations, obtained from burden distributions of the complete populations (blue) as well as sampled sets of cells (orange). c) VAF spectra measured in the complete population (blue) and a sampled set of cells (orange). A single simulation result is shown (diamonds) alongside the theoretically predicted expected value (Eq. (1)) (dashed line) and the average across 100 simulations (solid line). d) Distribution of NM and p inference results for 100 simulated and sampled populations, through estimation of and from the single cell burden distribution and fitting the number of lowest frequency (1/S) mutations to the theoretical prediction (Eq. (1)).

We apply our estimators to single-cell mutational data in blood obtained from the study by Lee-Six et al. [18], in which 89 haematopoietic stem cells (HSCs) were extracted from a bone marrow aspirate of a 59-year-old individual (Fig. 4a). Using Eq. (4) to estimate the expected number of divisions per lineage from the sample burden (SI 1.8.1), we find a total division rate (including both symmetric and asymmetric divisions) in the constant population phase of λ = 10.6 (7.6-15.3) divisions per HSC per year, which is within ranges previously suggested [3, 30]. Notably, the single-cell mutational burden alone can not disentangle symmetric and asymmetric division rates. Applying Eq. (5) to the sample burden distribution and taking Var(y)/E(y) = 1 (akin to assuming exponentially distributed division times, a common but simplified null model) we obtain an estimated mutation rate of per cell per division (Fig. 4a). This is significantly higher compared to the 1.2 mutations per division suggested previously [5, 18]. We posit different possible explanations for this discrepancy. Stem cell divisions and mutation accumulation are stochastic processes, therefore variation between individuals is expected. However, simulations of our model for a wide range of parameter values suggest the distribution of these inferred mutation rates is unlikely to present a wide enough variance to explain this difference (Fig. 7). Recently, Mitchell and colleagues [12] suggested that many mutations in somatic tissue are acquired independently of cell divisions. In fact, the previous estimates of 1.2 to 1.3 mutations per genome per division were derived from early developmental cell divisions where effects of aging would be negligible. The increased mutation rate inferred here could in part be explained by proliferation independent effects. However, it is important to note that the true distribution of divisions per lineage may also be over-dispersed compared to the Poisson model (i.e. Var(y) > E(y)), which would lead to an overestimation of µ in Eq. (5). Furthermore, the observed burden distribution is incompatible with a simple division-independent mutation model, which would lead to a Poisson distribution of the burdens that has a variance much smaller than what we observed (Fig. 4a). A combination of division independent mutations together with non-Poissonian cell divisions could possibly reproduce the result.

Evolutionary inferences in single cell HSC data

(a) The single-cell mutational burden distribution of the data (bars) and the Compound Poisson distribution obtained from its mean and variance, used to obtain the estimated per division mutation rate . (b) distribution of mutation frequencies of the data and theoretically predicted average fitted to only the lowest frequency (1/S) data point. (c) Difference Δvf between measured value of the VAF spectrum at lowest frequency (1/S) and its prediction from Eq. (1), for varying total population size N and asymmetric division proportion p, with fixed maturation time tM = 5 and operational hematopoietic population size NH = 50. The solid line denotes the plane of best fit where this difference is 0. (d) Maximally inferred population size N (taking p = 0 in (c)) for variation of the maturation time tN and the operational hematopoietic population size NH.

Comparison of an exponential and logistic growth model.

(a) Population size over time for exponential (dotted) and logistic (dashed) growth functions. At the point of maturity (t = 20) the exponential function reaches the population capacity (NM = 10′000), and the logistic function equals 0.99NM. The time points at which the VAF spectra are measured are indicated by solid vertical lines: t1 = 25 (blue) and t2 = 50 (orange). (b) Comparison of the expected VAF spectra, calculated with Eq. (1), in the exponential (dotted) and logistic (dashed) growth models measured at the time points t1 and t2. For reference, the theoretical predictions for an exponentially growing population without cell death and a constant population in the long time limit are shown as solid purple and green lines respectively.

Stochastic simulations of the Variance/Mean of the mutational burden distribution over time for a per cell division mutation rate of µ = 1.3 and varying stem cell population size N and asymmetric division probability p.Stochastic fluctuations are pronounced for small population size N and low asymmetric division probability p.

Likelihood of the Variance/Mean to be in the interval 3 < θ < 5 for a per division mutation rate of µ = 1.3.

Sparse sampling, single cell derived VAF spectra and evolutionary inferences

Current methods limit us to whole genome information of at most a few thousand single cells per tissue or tumour [22]. In many situations this will only constitute a small fraction of the underlying long term proliferating cell population. For example, although the exact number of HSCs remains unknown, most recent estimates suggest a population size of 105 cells [3, 18]. In tumours, this number can be as large as 1010 cells [31]. As we have shown in the previous paragraph, the single-cell mutational burden distribution allows certain inferences even if sampling is sparse. If we however construct the VAF spectrum from such a small sample, the distribution of variants will be significantly transformed with respect to that of the total population (Fig. 3c). Since Eq. (1) gives the expected curve measured from the total population, one must include a correction, which is given by the expectation of hypergeometric sampling (i.e. without replacement). This can be obtained through the transformation (see Supplement 1.6) with the VAFs observed in the sample and S the sample size. Furthermore, from stochastic simulations we note that the variance on the distribution increases with variant frequency f, making the lowest frequency state (i.e. 1/S) the best candidate for comparative fitting and evolutionary inferences (see Supplement Fig. 9).

Taking the mutation rate from Werner et al. [5] and the total division rate obtained from the burden distribution as constant, we explored solutions of Eq. (1) (sampled by Eq. (6)) for a wide range of realistic values in the remaining parameter space {N, p, tM, NH}. Comparing the lowest sampled frequency state (f = 1/89) with the data (Fig. 4b) we identify a curve in the space of the numbers of HSCs N and p (fraction of divisions that are asymmetric) where theoretically predicted VAF distributions are identical (Fig. 4c). Interestingly, this curve in N-p space naturally identifies a maximal stem cell population size capable of producing the data, which occurs when stem cells divide exclusively symmetrically (p = 0). Variations of tM (the time until maturity of the HSC population) and NH (the HSC pool size at which maintenance begins to occur) had only a small effect on the this inference (Fig. 4d). This is partly because mutations arising during these phases are more likely to be found at higher frequencies upon measurement [13, 32]. We find the data to be congruent with an adult HSC pool of at most 200’000-300’000 cells, depending on the exact timings of the maturation phase (Fig. 4d). This estimate agrees with the original study and other independent inferences based on population data [3, 18]. However, this upper bound corresponds to the case of exclusively symmetric stem cell divisions. Accounting for the possibility of mutating asymmetric divisions reduces the estimated stem cell pool. For example, the data would be consistent with 50’000 stem cells if 90% (p = 0.9) of all stem cell divisions are asymmetric. In principle, even smaller stem cell pool sizes are possible, though decreasing an order of magnitude would imply an extremely large portion (> 99%) of asymmetric HSC divisions. We note that in the data non-developmental branchings are observed, which immediately implies that not all HSC divisions can be asymmetric (p < 1) and thus a scenario of an extremely small population of only a handful of HSCs maintaining heamatopoisis can be rejected. With an orthogonal population based method, Watson and colleagues estimated 25’000 as a reasonable lower bound for the number of HSCs. Based on our inference, this would imply 0.95 as an upper bound for the proportion of all stem cell divisions that are asymmetric.

Discussion

Here, we have shown that single cell and bulk whole genome sequencing can inform on different aspects of somatic evolution. Single cell information does support previous bulk based estimates of possible ranges of HSC numbers. Surprisingly, the same single cell data leads to a mutation rate inference during homeostasis approximately 4 times higher compared to previous developmental estimates.

An important recent study has suggested an amended model of mutation accumulation in somatic tissues, wherein the majority of mutations are acquired continuously over time throughout a cell’s lifespan rather than during DNA replication [12]. We note that for a time dependent rate of mutation accumulation that is constant across all cells at a given point in time, this model predicts Poisson distributed single-cell mutational burdens. In contrast, we find the single-cell mutational burden distribution in human HSCs to be highly over-dispersed compared to a Poisson model. A time-associated model of mutation accumulation with a constant rate alone cannot generate such a pattern. At least one additional source of stochasticity would be needed. This could be fluctuations in the time dependent mutation rates, non-constant cell proliferation rates, e.g. bursts of stem cell divisions potentially in reaction to injury or disease, or a combination of both effects. Either way, our observations suggest that our current theoretical models of somatic evolution in healthy tissues are incomplete. There is evidence for unknown processes contributing to intra-tissue genetic heterogeneity. The precise nature of these processes remains an open question.

eLife Assessment

In this fundamental study, the authors propose analytical methods for inferring evolutionary parameters of interest from sequencing data in healthy tissue relevant to hematopoiesis. By combining analyses of single cell and bulk sequencing data, the authors can use a stochastic process to inform different aspects of genetic heterogeneity. The strength of evidence in support of the authors' claim is thus compelling. The work will be of broad interest to cell biologists and theoretical biologists.

Reviewer #1 (Public Review):

Summary:

Authors propose mathematical methods for inferring evolutionary parameters of interest from bulk/single cell sequencing data in healthy tissue and hematopoiesis. In general, the introduction is well-written and adequately references the relevant and important previous literature and findings in this field (e.g. the power laws for well-mixed exponentially growing populations). The authors consider 3 phases of human development: early development, growth and maintenance, and mature phase. In particular, time-dependent mutation rates in Figure 2d is an intriguing and strong result, and the process underlying Figures 3 and 4 are generally well-explained and convincing.

Notes & suggestions:

1. The explanation of Figure 2 in Lines 101 - 111 should be expanded for clarity. First, is Figure 2a derived from stochastic simulation (line 101 suggests) or some theoretical analysis? Second, the gradual transition from f-2 to f-1 is appreciated, but the shape of the intermediates is not addressed in detail. The power laws are straight lines, and the simulations provide curved lines -- please expand in what range (low or high frequency variants) the power law approximations apply.

Additionally, I do not understand the claim in line 108, that the transition is fast for low frequency variants, as the low frequency (on the left of the graph) lines are all close together, whereas the high frequency lines are far apart.

It would be helpful to reiterate in this paragraph that these power laws are derived based on exponentially growing populations and are expected to break down under homeostatic conditions.

2. The sample vs population (blue vs orange) in Figure 3 is under-explained. How is it that the mutational burden and inferred mutation rate in A and B roughly match, but the VAF distributions in C are so different? How was the sampled set chosen? Perhaps this is an unimportant distinction based on the particular sample set, but the divergence of the two in C may serve as a distraction, here.

3. The comparison of results herein to claims by Mitchell (ref. 12) are quite important results within the paper. I appreciate the note in the final paragraph of the discussion, and I suggest adding a sentence referencing the result noted in line 248-249 to the abstract, as well.

Reviewer #2 (Public Review):

Summary: The authors provide a nice summary on the possibility to study genetic heterogeneity and how to measure the dynamics of stem cells. By combining single cell and bulk sequencing analyses, they aim to use a stochastic process and inform on different aspects of genetic heterogeneity.

Strengths: Well designed study and strong methods

Weaknesses: Minor

Further clarification to Figure 3 legend would be good to explain the 'no association' of number of samples and mutational burden estimate as per line 180-182 p.8

Author Response

The following is the authors’ response to the original reviews.

We are very grateful for your time and efforts spent on our manuscript. Your feedback has been very valuable. Please see below a point-by-point response to each suggestion and actions taken to address each point in the manuscript.

eLife assessment

In this fundamental study, the authors propose analytical methods for inferring evolutionary parameters of interest from sequencing data in healthy tissue relevant to hematopoiesis. By combining analyses of single cell and bulk sequencing data, the authors can use a stochastic process to inform different aspects of genetic heterogeneity. The strength of evidence in support of the authors' claim is thus compelling. The work will be of broad interest to cell biologists and theoretical biologists.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

Authors propose mathematical methods for inferring evolutionary parameters of interest from bulk/single cell sequencing data in healthy tissue and hematopoiesis. In general, the introduction is well-written and adequately references the relevant and important previous literature and findings in this field (e.g. the power laws for well-mixed exponentially growing populations). The authors consider 3 phases of human development: early development, growth and maintenance, and mature phase. In particular, time-dependent mutation rates in Figure 2d is an intriguing and strong result, and the process underlying Figures 3 and 4 are generally wellexplained and convincing.

Thank you for your positive comments.

Notes & suggestions:

1. The explanation of Figure 2 in Lines 101 - 111 should be expanded for clarity. First, is Figure 2a derived from stochastic simulation (line 101 suggests) or some theoretical analysis? Second, the gradual transition from f-2 to f-1 is appreciated, but the shape of the intermediates is not addressed in detail. The power laws are straight lines, and the simulations provide curved lines -- please expand in what range (low or high frequency variants) the power law approximations apply.

Figure 2a was obtained from a numerical solution of equation 1, which describes the time dynamics of the expected VAF distribution. This is indeed unclear from the text, and we thank the reviewer for pointing out this discrepancy.

We thank the reviewer for this suggestion and have now adjusted this in the text (102-110):

“Numerical solutions of Eq.(1) show that the expected VAF distribution exhibits a gradual transition from the f-2 (growing population) to the f-1 (constant population) power law (Fig.2). These transitional states themselves do not adhere to some intermediate power-law (e.g. f for 1<<2), but instead present a sigmoidal shape, with the low frequency portion following f-1 and the high frequencies f-2 . Over time the shape changes as a wavelike front traveling from low to high frequency, with the constant-size equilibrium establishing earliest at the lowest frequencies and moving to higher frequency over time. Interestingly, the convergence towards equilibrium slows down over time -- for evenly-spaced observation times the solutions lie increasingly closer together -- further decreasing the speed at which the high frequency portion of the spectrum approaches equilibrium.”

We also changed the caption of Figure 2 to make this clearer as

“(a) Expected VAF distributions from evolving Eq1 to different time points for a population with an initial exponential growth phase and subsequent constant population phase (mature size N=103). Once the population reaches the maximum carrying capacity, the distribution moves from a 1/f2 growing population shape (purple) to a 1/f constant population shape (green). Note that the shift slows considerably at older age.”

In addition, we have also added annotations to Figure 2a and 2b to further clarify which line (green or purple) is f-1 and f-2.

Additionally, I do not understand the claim in line 108, that the transition is fast for low frequency variants, as the low frequency (on the left of the graph) lines are all close together, whereas the high frequency lines are far apart.

The lines are closer together in the low frequency portion (left of the plot) because they are already very close to the constant-size equilibrium (f-1/green line) and these frequencies approached equilibrium very fast. On the contrary, in the high frequency portion (right side of plot) they are still very far from equilibrium and approached equilibrium much slower.

It would be helpful to reiterate in this paragraph that these power laws are derived based on exponentially growing populations and are expected to break down under homeostatic conditions.

We have adjusted the relevant paragraph in the text to make the validity of the power laws clearer (90-94):

“For a well-mixed exponentially growing population without cell death the VAF spectrum 𝑣(𝑓) is given by 2𝜇/(𝑓 + 𝑓2 )$ (a 𝑓−2 power law) and is independent of time. In contrast, for a population of constant size – i.e. where birth and death rates are equal – the spectrum obeys 𝑣(𝑓) ∝ 2𝜇/ 𝑓 (a 𝑓−1 power law; see also SI), though this solution is only valid at sufficiently long times.”

1. The sample vs population (blue vs orange) in Figure 3 is under-explained. How is it that the mutational burden and inferred mutation rate in A and B roughly match, but the VAF distributions in C are so different? How was the sampled set chosen? Perhaps this is an unimportant distinction based on the particular sample set, but the divergence of the two in C may serve as a distraction, here.

This is an important question, and the answer was perhaps underemphasized in the caption. The sampling was performed as a uniform random sampling with replacement, and the same sample set was used for both the mutational burden and the VAF distribution. The reason for this stark contrast is that while the expectation of the burden distribution is not affected by sampling (i.e. sampling only affects the resolution/amount of stochasticity), the expectation of the VAF distribution changes due to sampling. While this was discussed in the section "Sparse sampling, single cell derived VAF spectra and evolutionary inferences", we have added note of this (indeed surprising) effect in the caption as well:

“(b) Distribution of estimated mutation rates from 10'000 individual simulations, obtained from burden distributions of the complete populations (blue) as well as sampled sets of cells (orange). Because the expected mutational burden distribution is unaltered by sampling, the expected estimate of the mutation rate from (5) remains unchanged: 𝐸(𝜇̃𝑝𝑜𝑝) = 𝐸(𝜇̃𝑠𝑎𝑚𝑝𝑙𝑒). However, sampling increases the noise on the observed burden distribution, which results in a higher errormargin of the estimate: 𝜎(𝜇̃𝑝𝑜𝑝) < 𝜎(𝜇̃𝑠𝑎𝑚𝑝𝑙𝑒).”

“(c) VAF spectra measured in the complete population (blue) and a sampled set of cells (orange). In contrast with the mutational burden distribution, strong sampling changes the shape of the expected distribution. A single simulation result is shown (diamonds) alongside the theoretically predicted expected values for both the total and sampled populations (Eqs. (1) and (6))(dashed line) and the average across 100 simulations (solid line).”

1. The comparison of results herein to claims by Mitchell (ref. 12) are quite important results within the paper. I appreciate the note in the final paragraph of the discussion, and I suggest adding a sentence referencing the result noted in line 248-249 to the abstract, as well.

We agree with the reviewer. We have extended the abstract now to reference the result in more detail:

“However, the single cell mutational burden distribution is over-dispersed compared to a model of Poisson distributed random mutations suggesting. A time-associated model of mutation accumulation with a constant rate alone cannot generate such a pattern. At least one additional source of stochasticity would be needed. Possible candidates for these processes may be occasional bursts of stem cell divisions, potentially in response to injury, or non-constant mutation rates either through environmental exposures or cell intrinsic variation.”

Reviewer #2 (Public Review):

Summary: The authors provide a nice summary on the possibility to study genetic heterogeneity and how to measure the dynamics of stem cells. By combining single cell and bulk sequencing analyses, they aim to use a stochastic process and inform on different aspects of genetic heterogeneity.

Strengths: Well designed study and strong methods

Thank you for your positive comments.

Weaknesses: Minor

Further clarification to Figure 3 legend would be good to explain the 'no association' of number of samples and mutational burden estimate as per line 180-182 p.8.

We have added a note to the caption of Figure 3b to explain more clearly how sampling affects the burden distribution and the mutation rate inferred from it (see also previous response to Reviewer 1):

“Because the expected mutational burden distribution is unaltered by sampling, the expected estimate of the mutation rate from (5) remains unchanged: 𝐸(𝜇̃𝑝𝑜𝑝) = 𝐸(𝜇̃𝑠𝑎𝑚𝑝𝑙𝑒). However, sampling increases the noise on the observed burden distribution, which results in a higher errormargin of the estimate: 𝜎(𝜇̃𝑝𝑜𝑝) < 𝜎(𝜇̃𝑠𝑎𝑚𝑝𝑙𝑒).”

Reviewer #1 (Recommendations For The Authors):

Minor/editorial suggestions:

1. Equation 1, please define \partial_t and \partial_K, for clarity.

These have now been defined in the text (between line 85-86): “where 𝜅 = 𝑓𝑁(𝑡) denotes the number of cells sharing a variant (the variant frequency f times the total population size N), 𝛿(x) is the Dirac impulse function, 𝜕𝑡 and 𝜕𝜅 are the partial derivatives with respect to time and variant size.”

1. Figure 2: It would be helpful to label the green and purple lines with the corresponding 1/f and 1/f^2 rule, in addition to the growing/fixed label, for clarity.

We agree and have now added the corresponding labels to each line.

Reviewer #2 (Recommendations For The Authors):

Minor suggestions are given below:

It would be nice for the authors to comment on whether the results could be extended/modified to account for possible fitness advantage of mutations which would be clinically relevant, for instance in the case of CHIP mutations and difference in time to myeloid malignancies transformation between CHIP/No CHIP individuals.

This is an important point. We agree with the reviewer that CHIP mutations play an important role in shaping mutational diversity especially in older individuals. Evidence is now emerging that CHIP mutations are almost universally present in individuals 60+. Interestingly, in individuals younger than 60, a neutral model (as presented here), does capture the observed effective dynamics well. For the purpose of the analysis underlying this manuscript, a neutral model seems reasonable.

The techniques we use here can be adjusted to include selection. How the results extend or modify will critically depend on the actual model of selection (rare or frequent CHIP mutations, strong vs weak selection etc.) that is realized in human hematopoiesis. Here we would say, the underlying biology currently is mostly unknown and is subject to (by others and in part by us) ongoing investigations, which extend beyond the scope of this manuscript.

We now make note of this point in the manuscript and added a small paragraph in page 11 to the discussion:

“Another open question is the role of selection and how it shapes intra-tissue genetic heterogeneity. Evidence is emerging that positively selected variants in blood are almost universally present in individuals above 60, while the effective observable dynamics in younger individuals is well described by neutral dynamics. How results presented here generalize or modify will critically depend on the model of selection realized in human hematopoiesis, e.g. a models of rare or frequent driver events. Details of the underlying biology are currently unknown.”

It would be nice to see if any significant differences in parameter estimates occur between loci with/without linkage disequilibrium, for instance HLA region. Could the number of single-cell samples be 'more' relevant when studying the VAF distribution in HLA region?

This is a good suggestion. We might be wrong or missing an important point, but somatic evolution as we use it in our modeling here is solely driven by asexual reproduction of cells. As such the entire genome of the cell is in linkage disequilibrium, independent of the precise genomic region (somatic evolution is in first approximation blind to germline mutations, as they are present in every single cell of the organism and therefore do not carry any information on the somatic evolutionary dynamics).

We thank all editors and reviewers again for your constructive comments.

Version 2

Abstract

Abstract

Intra-tissue genetic heterogeneity is universal to both healthy and cancerous tissues. It emerges from the stochastic accumulation of somatic mutations throughout development and homeostasis. By combining population genetics theory and genomic information, genetic heterogeneity can be exploited to infer tissue organisation and dynamics in vivo. However, many basic quantities, for example the dynamics of tissue specific stem cells remain difficult to quantify precisely. Here we show that single cell and bulk sequencing data inform on different aspects of the underlying stochastic processes. Bulk derived variant allele frequency spectra (VAF) show transitions from growing to constant stem cell populations with age in samples of healthy oesophagus epithelium. Single-cell mutational burden distributions allow a sample size independent measure of mutation and proliferation rates. Mutation rates in adult haematopietic stem cells are higher compared to inferences during development, suggesting additional proliferation independent effects. Furthermore, single cell derived VAF spectra contain information on the number of tissue specific stem cells. In haematopiesis, we find approximately 2×105 HSCs, if all stem cells divide symmetrically. However, the single cell mutational burden distribution is over-dispersed compared to a model of Poisson distributed random mutations suggesting. A time-associated model of mutation accumulation with a constant rate alone cannot generate such a pattern. At least one additional source of stochasticity would be needed. Possible candidates for these processes may be occasional bursts of stem cell divisions, potentially in response to injury, or non-constant mutation rates either through environmental exposures or cell intrinsic variation.

Introduction

Intra-tissue genetic heterogeneity emerges from a multitude of dynamical processes [1, 2]. Cells randomly accumulate mutations, they self-renew, differentiate or die and clones of varying fitness may compete and expand [3–6]. These core evolutionary principals form the basis for describing the aging of somatic tissues, tumour initiation and tumour progression [7–12]. Naturally, there is great interest in understanding precisely how these dynamical processes act. This typically involves hypothesizing models of cell behavior, and deriving quantitative estimates of their underlying physical parameters [3, 5, 13–17]. These could by quantities such as mutation rates per cell division, active stem cell numbers, symmetric and asymmetric division rates, etc.. Some such parameters are already well characterised in certain tissues. We know for example that during early development healthy somatic cells accumulate on average between 1.2 and 1.3 mutations per genome per division [5, 18]. Other parameters, such as the number of stem cells in a tissue type, are well known in some tissues but harder to quantify in others.

In general, a lack of time resolved information complicates evolutionary inferences, and we must often rely on indirect information, for example the patterns of tissue-specific genetic heterogeneity within and across individuals [19]. These observed patterns of genetic heterogeneity emerge from the underlying stochastic processes and can strongly depend on the measurement technique [20]. Both enforce specific limitations on the inferability of the underlying dynamics. Different types of genomic data – such as bulk and single cell whole genome sequencing – may contain different information on the system, and each comes with their own limitations. When bulk sequencing data is used for inference, some evolutionary parameters such as the population size and proliferation rates of stem cells are entangled and cannot be obtained in isolation [3, 16]. If on the other hand single cell data is employed, some additional information becomes available, for example the individual cell mutational burdens and the co-occurrence of mutations. However, often the sampling size is sparse compared to the presumed number of long-term proliferating cells in many tissues [18, 21, 22], which in turn introduces other constraints. Here we show that combining such data obtained from different resolutions (bulk or single cell) can help to overcome limitations and further narrow possible ranges of inferred parameters. Here, we develop concrete mathematical and computational models for extracting evolutionary parameters from bulk and single cell information and apply these methods to whole genome bulk sequencing data in healthy oesophagus [10] and whole genome single cell sequencing data in healthy haematopoiesis [18].

Results

A stochastic model of mutation accumulation in healthy somatic tissue

We model the stem cell dynamics of healthy tissue as a collection of individual cells that divide, differentiate, and die stochastically at predefined rates (see Fig.1). Novel mutations can occur with every cell division, each daughter acquiring a random number drawn from a Poisson distribution with rate μ. We explicitly include symmetric and asymmetric stem cell divisions at potentially different rates, the former resulting in two stem or two differentiated cells, and the latter in one stem and one differentiated cell. Since differentiated cells are lost from the stem cell pool they are functionally equivalent to cell death. While only symmetric divisions allow variants to change in frequency, both division types introduce novel variants into the population and thus both contribute to the evolving heterogeneity.

The distribution of variant allele frequencies changes with the growth phases and by sampling.

(a) In the current population, cells divide symmetrically into two daughter cells or asymmetrically with only one daughter cell kept in the focused population. All other events are mathematically equivalent to and are treated as a part of cell death. (b) The rates of symmetric and asymmetric division change during the population growth and lead to a dynamical distribution of variant allele frequencies. (c) The observed VAF distribution is shifted again during sampling compared to the VAF of the whole population – a fact should be considered when inferring population properties through genetic data.

Mutation accumulation occurs through all developmental stages from conception to death [23]. We therefore consider three phases of demographic changes of the stem cell population. In the first early developmental phase the stem cell population rapidly expands from a single cell exclusively by symmetric divisions at rate γ. This is followed by a growth and maintenance phase wherein the population continues to expand but also undergoes turnover through asymmetric divisions at rate ϕ, as well as cell removal (due to differentiation or death) and replacement at rate ρ. In the final mature phase, cell turnover continues but the size of the total stem population N remains constant.

We employ two independent implementations of our theoretical model. One is a direct stochastic simulation utilizing a Gillespie algorithm. This approach creates a simulated dataset containing all single cell or bulk sequencing properties, however it can be computationally expensive. In addition, we compute the time-dynamical expected value of the distribution of variant allele frequencies (the VAF spectrum) directly, which we find obeys the partial differential equation where κ = f N(t) denotes the number of cells sharing a variant (the variant frequency f times the total population size N), δ(x) is the Dirac impulse function, ∂t and ∂κ are the partial derivatives with respect to time and variant size, and This set of equations allows for computationally efficient numerical solutions (SI 1.3).

Transition of developmental growth and constant population size signatures in bulk whole genome sequencing data of healthy adult oesophagus

We first discuss established properties of the VAF spectrum relating to somatic tissues. In certain model systems it is given by a power law ∝ f −α with a critical exponent α. The value of the exponent depends on the demographic dynamics of the population. For a well-mixed exponentially growing population without cell death the VAF spectrum v(f) is given by 2μ/(f + f 2) (a f −2 power law) and is independent of time [6]. In contrast, for a population of constant size – i.e. where birth and death rates are equal – the spectrum obeys v(f) ∝ 2μ/ f [24] (a f −1 power law; see also SI 1.5.1), though this solution is only valid at sufficiently long times. In the following, we focus on the dynamics of genetic diversification in healthy tissues and show that the expected VAF spectrum contains both contributions of growing and constant population dynamics.

Healthy adult somatic tissues are thought to initially expand rapidly during ontogenic growth, continue to expand during infancy and childhood, and reach homeostasis by adulthood. It is thus natural to ask what the expected VAF spectrum would look like in tissues experiencing homeostasis after a period of growth. We first investigated the underlying VAF dynamics in a minimal theoretical model where the population switches from pure exponential growth to constant size once a maximal cell number is reached. Numerical solutions of Eq. (1) show that the expected VAF distribution exhibits a gradual transition from the f −2 (growing population) to the f −1 (constant population) power law (Fig. 2a). These transitional states themselves do not adhere to some intermediate power-law (e.g. f −α for 1 < α < 2), but instead present a sigmoidal shape, with the low frequency portion following f −1 and the high frequencies f −2. Over time the shape changes as a wavelike front traveling from low to high frequency, with the constant-size equilibrium establishing earliest at the lowest frequencies and moving to higher frequency over time. Interestingly, the convergence towards equilibrium slows down over time – for evenly-spaced observation times the solutions lie increasingly closer together – further decreasing the speed at which the high frequency portion of the spectrum approaches equilibrium. More complex forms of such demographic changes, for example the inclusion of a mixed growth-and- maintenance phase or a logistically growing population, result in qualitatively similar behavior (see Supplementary Fig. 5).

Bulk sequencing based VAF and mutation rate inferences in healthy esophagus.

(a) Expected VAF distributions from evolving Eq. (1) to different time points for a population with an initial exponential growth phase and subsequent constant population phase (mature size N = 103). Once the population reaches the maximum carrying capacity, the distribution moves from a 1/ f 2 growing population shape (purple) to a 1/ f constant population shape (green). Note that the shift slows considerably at older age. (b) VAF from healthy tissue in the oesophagus of 9 individuals sorted into age brackets. The youngest bracket, 20-39, is closer to the developmental 1/ f 2 scaling. The older age brackets are both close to the constant population 1/ f scaling, resembling the theoretical expectations.(c)Inferred mutation rates increase linearly with age. (d) Simulations of slowly growing stem cell populations reveal that mutation rates appear to increase with age, although the true underlying per division mutation rate remaining constant.

To test whether adult human tissues show such transitional signatures, we grouped the VAFs of healthy adult oesophagus samples from Martincorena and colleagues [10] into three age groups (young, middle and old, Fig. 2b). In accordance with our theoretical expectations, the averaged VAF spectrum of the young donor group is closest to the expected f −2 distribution of ontogenic growth, with low frequency variants only starting to approach the f −1 homeostatic scaling. The averaged spectra of the middle and old age groups on the other hand notably shift towards the expected f −1 homeostatic line. Interestingly, while there is a clear separation between the spectra of the youngest and middle age groups, those of the middle and older groups are much closer. This agrees with our prediction that the speed of convergence of the spectra towards homeostatic scaling slows down with age.

VAF based mutation rate inferences in healthy tissues

The theoretical prediction for the VAF spectrum of an exponentially growing population, which is recovered in many cancers given sufficient sequencing depth [20], contains information on the effective mutation rate μ [25]. Similarly, the VAF spectrum of a constant population includes the same mutation rate μ (SI). Thus we can in principle infer μ from the previously described oesophagus data, for example by applying a regression approach. However, since the scaling of the VAF spectrum of healthy tissues can be shown to be age dependent (Fig. 2a-b), we use two reference shapes for a growing and a constant population, respectively. The fitting is performed by a linear regression least-squares approach. The general solution is computed from the VAF spectrum of the data, Vd(μ), and the reference shape Vr, which assumes μ = 1, as follows: If Vd(μ) has the shape of Vr, then E(Vd(μ)) = μ E(Vr), which can be used to prove equation 2: The VAF shapes for the growing and constant population give close estimates, as can be seen in fig. 2, which compares the age of all individuals with corresponding inferred effective mutation rates. The estimates are in the range of 1 to 2 mutations per genome per cell division and agree with previous observations based on early developmental stem cell divisions [18]. Estimates based on a growing population are slightly higher compared to estimates based on a constant population, but differences are small.

Surprisingly, we observe a clear trend of a linearly increasing effective mutation rate with age. This trend cannot be explained by the transition of variants from a growing into a constant population alone. An actual increase of the true mutation rate per cell division with age would be a direct explanation. However, this seems unlikely: It is well established that the mutational burden across individuals and tissues increases linearly with age [9, 11, 26], and an increasing mutation rate would in contrast result in an accelerated increase of mutational burden with age, which is not observed experimentally.

An alternative explanation is a continued slow linear increase of the stem cell population with age (Fig. 2d). In such a scenario, the effective mutation rate would appear to increase linearly with age as well, although the true mutation rate per cell division remains unchanged. At first glance, a linearly increasing stem cell population appears unnatural. However, such a linear increase would emerge from a small bias of stem cells towards self-renewal (growth) and concomitantly a slowdown of the stem cell proliferation rate proportional to the change of population size [27]. The latter is a natural feedback mechanism ensuring a constant output of differentiated somatic cells. Direct in vivo imaging in multiple human tissues including Oesaphagus seem to support an increased density and slower proliferation of stem cells with age [28]. This explanation is attractive for another reason: A feedback of decreased cell proliferation with increasing population size would also function as a tumour suppressor mechanism, and may be another reason cancer driver mutations in healthy tissues are abundant but rarely progress to cancer [10, 29]. Progression would require additional genomic or environmental changes to overcome this regulatory feedback.

Single-cell mutational burden allows sample size independent inferences of stem cell proliferation and mutation rates

With the increasing availability of single-cell data, it becomes possible to investigate distributions of certain quantities for which bulk sequencing only provides average measurements. One such is the number of mutations present in individual cells. Recent work by Lee-Six and colleagues [18] showed that in a 59 year old donor haematopoietic stem cells had accumulated on average around 1000 mutations per cell; however there was significant variation between individual genomes, some carrying 900 mutations, while others up to 1200. This variation is the result of the stochastic nature of mutation accumulation and cell division, and can be exploited to estimate per division mutation and stem cell proliferation rates. To this end we look at the distribution of the number of mutations per cell, which we will from here on refer to as the mutational burden distribution. Formally, we describe the mutational burden mj in a cell j by a stochastic sum over its past divisions , in which both the total number of divisions yj in the cell’s past and the number of mutations per division uij are random variables. Even without knowing the actual distributions of y and u, general expressions for the expectation and variance of mj that only depend on the expectation and variance of y and u exist and are given by E(m) = E(y) E(u) and Var(m) = E(y) Var(u) + E(u)2 Var(y) (SI 1.7). These expressions allow direct estimates of the effective division and mutation rate. Assuming the mutation rate per daughter cell is Poisson distributed with expectation μ, the average number of divisions per lineage — expressed in terms of an effective division rate λ — becomes In addition, if the mutation rate μ is unknown, it can similarly be obtained from We first show that these estimators recover per division mutation and proliferation rates in stochastic simulations (Fig. 3a and b). More importantly, they only require information on comparably few single cells. Approximately, 100 cells are sufficient to reliably reconstruct the mutational burden distribution, which for example would constitute a small sample of the haematopoietic stem cell pool. In fact, the inference does not significantly improve if we increase data resolution beyond a few hundred single cells (Fig. 3a and b). This is in stark contrast to inferences based on single cell phylogeny: Although they contain more information in principle, important aspects of a single cell phylogeny, e.g. the number of observable branchings, are strongly affected by sampling and stem cell population size.

Inference of evolutionary parameters on simulated stem cell populations.

Simulated populations were run up to age 59, growing exponentially from a single cell to constant size NM = 10′000 at age tM = 5, with mutation rate μ = 1.2 and division rates λ = 5 and p = 0.4. Where sampling is mentioned, the sample size 89 was taken. a) The single-cell mutational burden distribution. The compound Poisson distribution (dashed line) matches the burden distribution when averaging over multiple independently evolved populations (filled curve). b) Distribution of estimated mutation rates from 10’000 individual simulations, obtained from burden distributions of the complete populations (blue) as well as sampled sets of cells (orange). Because the expected mutational burden distribution is unaltered by sampling, the expected estimate of the mutation rate from Eq. (5) remains unchanged: . However, sampling increases the noise on the observed burden distribution, which results in a higher error-margin of the estimate: . c) VAF spectra measured in the complete population (blue) and a sampled set of cells (orange). In contrast with the mutational burden distribution, strong sampling changes the shape of the expected distribution. A single simulation result is shown (diamonds) alongside the theoretically predicted expected values for both the total and sampled populations (Eqs. (1) and (6)) (dashed line) and the average across 100 simulations (solid line). d) Distribution of NM and p inference results for 100 simulated and sampled populations, through estimation of and from the single cell burden distribution and fitting the number of lowest frequency (1/S) mutations to the theoretical prediction (Eq. (1)).

We apply our estimators to single-cell mutational data in blood obtained from the study by Lee-Six et al. [18], in which 89 haematopoietic stem cells (HSCs) were extracted from a bone marrow aspirate of a 59-year-old individual (Fig. 4a). Using Eq. (4) to estimate the expected number of divisions per lineage from the sample burden (SI 1.8.1), we find a total division rate (including both symmetric and asymmetric divisions) in the constant population phase of λ = 10.6 (7.6-15.3) divisions per HSC per year, which is within ranges previously suggested [3, 30]. Notably, the single-cell mutational burden alone can not disentangle symmetric and asymmetric division rates. Applying Eq. (5) to the sample burden distribution and taking Var(y)/ E(y) = 1 (akin to assuming exponentially distributed division times, a common but simplified null model) we obtain an estimated mutation rate of per cell per division (Fig. 4a). This is significantly higher compared to the 1.2 mutations per division suggested previously [5, 18]. We posit different possible explanations for this discrepancy. Stem cell divisions and mutation accumulation are stochastic processes, therefore variation between individuals is expected. However, simulations of our model for a wide range of parameter values suggest the distribution of these inferred mutation rates is unlikely to present a wide enough variance to explain this difference (Fig. 7). Recently, Mitchell and colleagues [12] suggested that many mutations in somatic tissue are acquired independently of cell divisions. In fact, the previous estimates of 1.2 to 1.3 mutations per genome per division were derived from early developmental cell divisions where effects of aging would be negligible. The increased mutation rate inferred here could in part be explained by proliferation independent effects. However, it is important to note that the true distribution of divisions per lineage may also be over-dispersed compared to the Poisson model (i.e. Var(y) > E(y)), which would lead to an overestimation of μ in Eq. (5). Furthermore, the observed burden distribution is incompatible with a simple division-independent mutation model, which would lead to a Poisson distribution of the burdens that has a variance much smaller than what we observed (Fig. 4a). A combination of division independent mutations together with non-Poissonian cell divisions could possibly reproduce the result.

Evolutionary inferences in single cell HSC data

(a) The single-cell mutational burden distribution of the data (bars) and the Compound Poisson distribution obtained from its mean and variance, used to obtain the estimated per division mutation rate . (b) distribution of mutation frequencies of the data and theoretically predicted average fitted to only the lowest frequency (1/S) data point. (c) Difference Δvf between measured value of the VAF spectrum at lowest frequency (1/S) and its prediction from Eq. (1), for varying total population size N and asymmetric division proportion p, with fixed maturation time tM = 5 and operational hematopoietic population size NH = 50. The solid line denotes the plane of best fit where this difference is 0. (d) Maximally inferred population size N (taking p = 0 in (c)) for variation of the maturation time tN and the operational hematopoietic population size NH.

Comparison of an exponential and logistic growth model.

(a) Population size over time for exponential (dotted) and logistic (dashed) growth functions. At the point of maturity (t = 20) the exponential function reaches the population capacity (NM = 10′000), and the logistic function equals 0.99NM. The time points at which the VAF spectra are measured are indicated by solid vertical lines: t1 = 25 (blue) and t2 = 50 (orange). (b) Comparison of the expected VAF spectra, calculated with Eq. (1), in the exponential (dotted) and logistic (dashed) growth models measured at the time points t1 and t2. For reference, the theoretical predictions for an exponentially growing population without cell death and a constant population in the long time limit are shown as solid purple and green lines respectively.

Stochastic simulations of the Variance/Mean of the mutational burden distribution over time for a per cell division mutation rate of μ = 1.3 and varying stem cell population size N and asymmetric division probability p.Stochastic fluctuations are pronounced for small population size N and low asymmetric division probability p.

Likelihood of the Variance/Mean to be in the interval 3 < θ < 5 for a per division mutation rate of μ = 1.3.

Likelihood of the Variance/Mean to be in the interval 3 < θ < 5 for a per division mutation rate of μ = 3.

Sparse sampling, single cell derived VAF spectra and evolutionary inferences

Current methods limit us to whole genome information of at most a few thousand single cells per tissue or tumour [22]. In many situations this will only constitute a small fraction of the underlying long term proliferating cell population. For example, although the exact number of HSCs remains unknown, most recent estimates suggest a population size of 105 cells [3, 18]. In tumours, this number can be as large as 1010 cells [31]. As we have shown in the previous paragraph, the single-cell mutational burden distribution allows certain inferences even if sampling is sparse. If we however construct the VAF spectrum from such a small sample, the distribution of variants will be significantly transformed with respect to that of the total population (Fig. 3c). Since Eq. (1) gives the expected curve measured from the total population, one must include a correction, which is given by the expectation of hypergeometric sampling (i.e. without replacement). This can be obtained through the transformation (see Supplement 1.6) with the VAFs observed in the sample and S the sample size. Furthermore, from stochastic simulations we note that the variance on the distribution increases with variant frequency f, making the lowest frequency state (i.e. 1/S) the best candidate for comparative fitting and evolutionary inferences (see Supplement Fig. 9).

The standard deviation on the VAF spectrum increases for higher frequencies.

a) The VAF spectrum Vf averaged across 100 simulations of a population evolved according to the model described in section 1.1. The standard deviation from the mean is shown as the band Vf ±σf around the average.(b)The standard deviation across all simulations for each frequency f = 1/N, 2/N, …, scaled by the average spectrum Vf.

Taking the mutation rate from Werner et al. [5] and the total division rate obtained from the burden distribution as constant, we explored solutions of Eq. (1) (sampled by Eq. (6)) for a wide range of realistic values in the remaining parameter space {N, p, tM, NH }. Comparing the lowest sampled frequency state (f = 1/89) with the data (Fig. 4b) we identify a curve in the space of the numbers of HSCs N and p (fraction of divisions that are asymmetric) where theoretically predicted VAF distributions are identical (Fig. 4c). Interestingly, this curve in N-p space naturally identifies a maximal stem cell population size capable of producing the data, which occurs when stem cells divide exclusively symmetrically (p = 0). Variations of tM (the time until maturity of the HSC population) and NH (the HSC pool size at which maintenance begins to occur) had only a small effect on the this inference (Fig. 4d). This is partly because mutations arising during these phases are more likely to be found at higher frequencies upon measurement [13, 32]. We find the data to be congruent with an adult HSC pool of at most 200’000-300’000 cells, depending on the exact timings of the maturation phase (Fig. 4d). This estimate agrees with the original study and other independent inferences based on population data [3, 18]. However, this upper bound corresponds to the case of exclusively symmetric stem cell divisions. Accounting for the possibility of mutating asymmetric divisions reduces the estimated stem cell pool. For example, the data would be consistent with 50’000 stem cells if 90% (p = 0.9) of all stem cell divisions are asymmetric. In principle, even smaller stem cell pool sizes are possible, though decreasing an order of magnitude would imply an extremely large portion (> 99%) of asymmetric HSC divisions. We note that in the data non-developmental branchings are observed, which immediately implies that not all HSC divisions can be asymmetric (p < 1) and thus a scenario of an extremely small population of only a handful of HSCs maintaining heamatopoisis can be rejected. With an orthogonal population based method, Watson and colleagues estimated 25’000 as a reasonable lower bound for the number of HSCs. Based on our inference, this would imply 0.95 as an upper bound for the proportion of all stem cell divisions that are asymmetric.

Discussion

Here, we have shown that single cell and bulk whole genome sequencing can inform on different aspects of somatic evolution. Single cell information does support previous bulk based estimates of possible ranges of HSC numbers. Surprisingly, the same single cell data leads to a mutation rate inference during homeostasis approximately 4 times higher compared to previous developmental estimates.

Another open question is the role of selection and how it shapes intra-tissue genetic heterogeneity. Evidence is emerging that positively selected variants in blood are almost universally present in individuals above 60, while the effective observable dynamics in younger individuals is well described by neutral dynamics. How results presented here generalize or modify will critically depend on the model of selection realized in human hematopoiesis, e.g. a models of rare or frequent driver events. Details of the underlying biology are currently unknown.

An important recent study has suggested an amended model of mutation accumulation in somatic tissues, wherein the majority of mutations are acquired continuously over time throughout a cell’s lifespan rather than during DNA replication [12]. We note that for a time dependent rate of mutation accumulation that is constant across all cells at a given point in time, this model predicts Poisson distributed single-cell mutational burdens. In contrast, we find the single-cell mutational burden distribution in human HSCs to be highly over-dispersed compared to a Poisson model. A time-associated model of mutation accumulation with a constant rate alone cannot generate such a pattern. At least one additional source of stochasticity would be needed. This could be fluctuations in the time dependent mutation rates, non-constant cell proliferation rates, e.g. bursts of stem cell divisions potentially in reaction to injury or disease, or a combination of both effects. Either way, our observations suggest that our current theoretical models of somatic evolution in healthy tissues are incomplete. There is evidence for unknown processes contributing to intra-tissue genetic heterogeneity. The precise nature of these processes remains an open question.